A Linear Solution of the Linearity Challenge

### 2025-12-18 paper/related.tex
related

\Cref{tab:comparison} summarizes a comparison among the formalizations of binary
session calculus that we are aware of, roughly sorted by publication date. We
have selected those formalizations that define a calculus of binary sessions
equipped with a reduction/evaluation semantics with proofs of subject reduction
and possibly of deadlock freedom.
%
The formalizations differ substantially in terms of reference calculus, used
proof assistant, process/expression/channel representation, handling of
linearity.
%
Most importantly, almost all of them predates the publlication of the linearity
challenge~\cite{CarboneEtAl24} hence were not necessarily targeted to solving
that particular problem. Nonetheless, they provide a broad spectrum of
techniques

\fullcite{GotoEtAl16} describe the oldest formalization of a session-based
calculus for which we were able to access the source code (the formalization of
\citet{Gay01} predates this work, but the source code is no longer publicly
available).

fullcite{RouvoetPoulsenKrebbersVisser20} also formalize a subset of \CP as case
study for their approach based on separation logic, which we have partially
taken inspiration from for the representation of processes
(\cref{sec:context-agda,sec:process-agda}). Their formalization adopts the same
intrinsically-typed representation of processes with nameless channels as our
own, but their formalization focuses on the implementation of an interpreter (no
small-step semantics is defined) and subject reduction is

@@

    \footnotesize

@@

    \begin{tabular}{@{}r@{~}|@{~}c@{~}|@{~}c@{~~}c@{~~}c@{~~}c@{~~}c@{~~}c@{~~}c@{~}|@{~}r@{~~}r@{~}|@{~}r@{~~}r@{~}|@{~}r@{~}l@{}}

@@

        \textlabel{Shared channels} &
        \textlabel{Polymorphism} &

@@

        \fullcite{GotoEtAl16} &

@@

@@

        \footnote{Excluding safety results.}

@@

        \fullcite{Thiemann19} &

@@

        \SessionLAM &

@@

@@

        \footnote{Including support for recursive types, subtyping and the interpreter.}

@@

        \fullcite{CastroFerreiraYoshida20} &
        \footnote{Excluding the original syntax.}
        \fullcite{CicconePadovani20} &

@@

@@

        \footnote{Including support for recursive and dependent session types.}
        \fullcite{RouvoetPoulsenKrebbersVisser20} &

@@

        \fullcite{ZalakainDardha21} &

@@

        \GradedPI &

@@

@@

        \footnote{Including library for algebra of types.}

@@

        \fullcite{JacobsBalzerKrebbers22} &

@@

@@

        \fullcite{SanoKavanaghPientka23} &

@@

        \fullcite{ZackonEtAl25} &

@@

        \footnote{Including deadlock freedom and excluding safety results.}

@@

>>>>

\Cref{tab:comparison} summarizes a comparison among the formalizations of binary
session calculi and of the linear $\pi$-calculus, sorted by publication date. We
have selected those formalizations that include the following characteristics:
\begin{itemize}
\item syntax and reduction/evaluation semantics for a calculus with support for
    binary sessions, either natively or through their encoding in terms of
    linear channels~\cite{KobayashiPierceTurner99,Kobayashi02b};
\item a type system along with the usual typing preservation results;
\item a proof of global progress/deadlock freedom, for those type systems that
    enforce it.
\end{itemize}

The formalizations differ substantially in terms of reference calculus, used
proof assistant, process/expression/channel representation, handling of
linearity.
%
Most importantly, almost all of them predate the publication of the linearity
challenge~\citep{CarboneEtAl24}, hence they are not necessarily targeted to
solving that particular problem.

\citet{GotoEtAl16} describe the formalization of a session-based variant of the
$\pi$-calculus which supports channel polymorphism. This is the oldest
formalization of a session-based calculus for which we were able to retrieve the
source code (the work of \citet{Gay01} predates this one, but its source code is
not publicly available any more).

\citet{Thiemann19} formalizes a subset of \GV~\citep{GayVasconcelos10}, a
functional language extended with session communication primitives, along with
an interpreter.

\citet{RouvoetPoulsenKrebbersVisser20} present a library of abstractions
inspired to separation logic aiding the formalization of interpreters for
computing models supported by linear type systems. One of the presented case
studies is the formalization of a fragment of \GV~\citep{Wadler14}.

% as case study for their approach based on separation logic, which we have
% partially taken inspiration from for the representation of processes
% (\cref{sec:context-agda,sec:process-agda}). Their formalization adopts the same
% intrinsically-typed representation of processes with nameless channels as our
% own, but their formalization focuses on the implementation of an interpreter (no
% small-step semantics is defined) and subject reduction is

\citet{CastroFerreiraYoshida20} describe EMTST, a library for the formalization
of session type systems which includes as case studies the session calculus of
\citet{HondaVasconcelosKubo98} (called ``original system'') and a revised
version of it that is more amenable to be formalized using a locally nameless
representation of channels.

\citet{CicconePadovani20} formalize a variant of the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99} that supports dependent types, so
as to enable the description of communication protocols whose structure depends
on the content of exchanged messages.

All the formalizations mentioned make use of context splitting. In contrast,
\citet{ZalakainDardha21} formalize a generalization of the linear $\pi$-calculus
which is parametric in a \emph{usage algebra} (to account for channel
sharing/linearity) and that is based on \emph{leftover typing}~\cite{}. The
basic idea is that a process is typed with respect to an \emph{input context},
which describes the overall available channels, and an \emph{output context},
which describes the remaining channels not consumed by the process. In this way,
it is possible to ``concatenate'' typing judgments instead of splitting them
beforehand.

Another work based on context splitting is the one by
\citet{JacobsBalzerKrebbers22}, who formalize a library of \emph{connectivity
graphs} for reasoning and enforcing deadlock freedom in a variant of
\GV~\citep{GayVasconcelos10,Wadler14}. This is the first formalization of
deadlock freedom for a calculus of sessions.

Motivated by the technical difficulties arising from the splitting of typing
contexts, \citet{SanoKavanaghPientka23} propose an alternative approach based on
\emph{linearity predicates}. The key idea is to treat typing contexts
structurally, and to enforce the linear usage of channels by means of separate
predicates checking the syntactical occurrence of channels.

\citet{ZackonEtAl25} describe yet another approach to the handling of typing
contexts whereby channels are associated not just with a type but also with a
\emph{tag} (or \emph{multiplicity}) from a given resource algebra for describing
their availability. This approach streamlines context splitting in the sense
that contexts can be treated in an essentially structural way, except for tags
which are conveninently combined using operations from the resource algebra.

@@

@@

    \begin{tabular}{@{}r@{~}|@{~}c@{~}|@{~}c@{~~}c@{~~}c@{~~}c@{~~}c@{~}|@{~}r@{~~}r@{~}|@{~}r@{~~}r@{~}|@{~}r@{~}l@{}}

@@

        % \textlabel{Shared channels} &
        % \textlabel{Polymorphism} &

@@

        \citet{GotoEtAl16} &

@@

        % \YES &
        % \YES &

@@

        \footnote{Including shared and polymorphic channels, excluding safety results.}

@@

        \citet{Thiemann19} &

@@

        \GV &

@@

        % \YES &
        % \NO &

@@

        \footnote{Including shared channels, recursive types, subtyping and the interpreter.}

@@

        \citet{RouvoetPoulsenKrebbersVisser20} &

@@

        \GV &

@@

        % \NO &
        % \NO &

@@

        \citet{CastroFerreiraYoshida20} &
        % \YES &
        % \NO &
        \footnote{Including shared channels, excluding the original syntax.}
        \citet{CicconePadovani20} &

@@

        % \YES &
        % \NO &

@@

        \footnote{Including shared channels, recursive and dependent session types.}
        \citet{ZalakainDardha21} &

@@

        \footnote{Including shared channels and the library for algebra of types.}

@@

        \citet{JacobsBalzerKrebbers22} &

@@

        % \NO &
        % \NO &

@@

        \citet{SanoKavanaghPientka23} &

@@

        \citet{ZackonEtAl25} &

@@

        \footnote{Including shared, polymorphic channels and deadlock freedom, excluding safety results.}

@@

instead of counting line numbers, which may be affected by indentation habits,
readability issues, and other syntactical restrictions of the proof assistant
being used, we count characters after spaces have been
squeezed.\footnote{Sequences of two or more consecutive space-like characters
are collapsed into a single space. The squeezing is obtained by running the
command \texttt{tr -s [:space:] file} on Unix-like systems.}
